---
title: "STAT240 Project: Titanic fare prices affect of passanger survivabililty"
output: html_document
date: "2026-04-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include= FALSE}
## Data Read & Clean

# For the group: My data was put into a separate data folder
data <- read.csv("../data/titanic.csv")

head(data) # Reads good!
```

```{r include= FALSE}
library(dplyr)
fare_survivability <- data %>% select("Survived", "Fare") # This keeps only the variables we need

# We will use these columns in the data
head(fare_survivability)

# Data Clean (remove NA's)
library(tidyr) # For drop_na

fare_survivability_cleaned <- fare_survivability %>% drop_na(Survived, Fare)
fare_survivability_cleaned
```


```{r include= FALSE}
# Actual T-test (For p-value)

# t-test shows p-value = 2.699e-11
t.test(Fare ~ Survived, data = fare_survivability_cleaned)
```


1.Jayden Merrick,
2.Samantha Laderman,
3.Dongwoon Han,
4.Gavin Ryan.
5.Hyewon Aha



### Introduction: 
The Titanic Disaster has been studied through many different factors, including class, gender, age, and access to lifeboats. In this project, we focus on whether ticket fare is related to passenger survival. Since fare price can reflect a passenger’s social class and location on the ship, comparing fare groups may help explain differences in survival outcomes.

Our question of interest is: Were people who paid higher fares more likely to survive the crash than people who paid lower fares?

The voyage of the Titanic saw a wide range of people from different social classes. There were approximately 324 first class passengers on board, 284 second class and 709 third class passengers. Tickets for a first class suite would cost around $133,132 today. Third class would cost around $1,071. 

The Titanic was equipped with only 20 lifeboats that could hold up to 1,178 people (far fewer than the number of people on board). Around 1,516 people perished due to the frigid waters and violence of the boat sinking. Of these 1,516 people, 818 passengers and 698 crew members died. 

Our results show that there is a direct correlation between fare and survivorship. The answer to our question (were people who paid higher fares more likely to survive the crash than people who paid lower fares) is that people who paid higher fares were more likely to survive the crash. 




### Background

For our project, we decided to utilize the Titanic database published by researcher Xiaming Chen on GitHub to examine whether passenger fare was related to survival. The dataset can be found on awesome-public-datasets/Datasets/titanic.csv.zip at master · awesomedata/awesome-public-datasets. Within the dataset, there are several columns of data describing passenger details. They are Passenger, Survived, Pclass, Name, Sex, Age, Sibsp, Parch, Ticket, Fare, Cabin, and Embarked. The Passenger column represents the number the passenger was assigned for identification in the dataset. Survived is represented in binary terms with a value of zero representing they did not survive and a value of one indicating survival. Pclass has values ranging from 1-3 representing the class level of the passenger with one representing first class, two representing second class, and three representing third class. Name is a string value of each passenger’s name on their respective boarding ticket. Sex is a binary string of either male or female. Sibsp is the count of siblings/spouses onboard and Parch is the count of parents/children onboard. Fare represents the price of each passenger’s ticket. The cabin represents the room number of each passenger. Embarked represents where each passenger got on the ship with C representing Cherbourg, Q representing Queenstown, and S representing Southampton.
For our purposes, we focused on the Survived and Fare columns to answer whether fare played a significant role in the odds of survival. To best achieve this, we used the ggplot2 library’s boxplot function to make a graph to display the data. Furthermore, we ran a standard t-test in R to get the p-value of $2.699e^{-11}$.



```{r echo = FALSE}
## BoxPlot Graph

library(ggplot2)

ggplot(fare_survivability_cleaned, aes(x = factor(Survived), y = Fare), fill = factor(Survived)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 300)) + #needed to zoom in due to outlier
  scale_x_discrete(labels = c("0" = "Died", "1" = "Survived")) + #Relable the 1 and 0's
  labs(
    title = "Comparing Fare rates of the Dead and Survived Titanic passengers",
    x = "Survival Status",
    y = "Fare Price"
  )


```

Fig 1. Boxplot graph shows two columns representing the binary states of the recorded fate of each person in our data, Died or Survived, against the fare rate paid by each individual. The graph shows a difference between the two columns, where individuals labeled as Survived have an Interquartile Range including higher fare prices.

The boxplot shows that passengers who survived generally paid higher fares than those who died. The median fare and interquartile range for survivors are both noticeably higher, indicating that a larger portion of survivors were in higher fare ranges. Although there is some overlap, the overall distribution is shifted upward for survivors. This visual difference suggests a possible relationship between ticket fare and survival status. However, a statistical test is needed to determine whether this observed difference is statistically significant


While the dataset includes many variables, our analysis focuses on the fare price and survival status because fare can reflect a passenger’s class and location on the ship. Passengers in higher classes were generally located closer to the deck and lifeboats, which may have increased their chances of survival. Therefore, fare is an important variable to study when examining differences in survival outcomes.

### Analysis


$$Hypotheses$$

$$H_0 = (\mu_1 = \mu_0) \implies (\mu_1 - \mu_0 = 0)$$
	
$$H_A = (\mu_1 > \mu_0) \implies (\mu_1 - \mu_0 > 0)~or~(\mu_1 - \mu_0 < 0)$$

Where $\mu_1$ is the mean fare for the survivors, and $\mu_0$$is the mean fare for passengers who did not survive.

The null hypothesis for our study would infer that there is no correlation between the fare price paid and survivability. Our alternative hypothesis would say the opposite, that there is a correlation between fare price and survivability where the higher the fare price increases the rate of survival.

<br>
<br>

$$Parameters~of~Interest$$
Survival (1 = Alive,0 = Died)

Fare rate (Range of \$1 to \$300)

$\mu_0$, the mean of dead individuals fare prices

$\mu_1$ the mean of surviving individuals fare prices


<br>
<br>

$$Type~of~Inference$$
We used a standard t-test for an inference of two population proportions.

<br>
<br>

$$Assumptions$$

Note: Notation is based off of Survival (1 = Survived, 0 = Died)

$$F_0 = \{Survived = 0\} \sim D_0(\mu_0, \sigma_0)$$
$$F_1 = \{Survived = 1\} \sim D_1(\mu_1, \sigma_1)$$


Normal distribution: Our t-test assumes Normal distribution despite our data likely not having a bell curve. However, this can still be used in the t-test due to the central limit theorem.

Independence: It is assumed that all fare prices were not based on each other for each of the 891 individuals.
 
<br> 
<br>
      
$$Test~Statistic$$
 
We used the t test question:
$$t = \frac{\bar{x}_1 - \bar{x}_0}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_0^2}{n_0}}}$$ 
 
The degrees of freedom were calculated by:
$df = \frac{\left( \frac{s_1^2}{n_1} + \frac{s_0^2}{n_0} \right)^2}{\frac{\left( \frac{s_1^2}{n_1} \right)^2}{n_1 - 1} + \frac{\left( \frac{s_0^2}{n_0} \right)^2}{n_0 - 1}}$$.

We got the degrees of freedom to be 436.7

### Results
P = $2.699e^{-11}$

Df = 436.7

T = -6.8391


### Discussion 

$$Further~Interpretation$$
We reject the null hypothesis as we have concluded that there is enough evidence that passengers who paid higher fares were more likely to survive. P-values are one way to determine whether or not to reject the null hypothesis. A large p-value lends support to a correct null hypothesis. Conversely, a small p-value means that there is a lesser chance that the data supports the null hypothesis. Our p-value is approximately  $2.699e^{-11}$. The p-value (probability value) is a variable which represents the probability of the result occurring assuming the null hypothesis is true. The higher the p-value, the more likely it is that we have got our result by mere chance, and not actual correlation between variables. Since this value is far below the standard significance level $(p < 0.5)$, it provides extremely strong evidence to reject the null hypothesis in favor of the alternative; that there is a correlation between fare price and survivability.

$$Shortcomings~of~the~analysis$$
One shortcoming of our dataset is that it didn’t encompass all passengers onboard the Titanic. There were 2,229 passengers, but our dataset only includes 891 people, which may lead to a sampling bias. It could be the case that our sample included more survivors that paid a higher fare price, which would have affected our results.
Another is possible confounding variables. Survival was not merely a matter of the demographics present in our data. Several critical factors that likely played a decisive role in saving lives were not captured. For instance, while we know lower fares are often correlated with being farther down in the boat, we lack specific data on the physical proximity of cabins to lifeboat stations and the specific evacuation protocols led by the crew on different decks. Additionally, human elements such as social ties (traveling with family vs alone) and the exact timing of when a passenger reached the boat deck are missing. Without these variables, our analysis may oversimplify the complex reality of the evacuation process, as these logistical and situational factors could have outweighed or significantly modified the influence of age and gender.

$$Additional~Questions/Further~Recommendations$$
One other question we had about this dataset was about gender and age. We wondered if there was a significantly higher proportion of female as well as child survivors as women and children were prioritized when it came to lifeboats. We can do a t-test to determine if these survival rate differences are statistically significant across all passenger classes.

We recommend expanding the dataset to include all passengers in future analyses. 
By running comparative tests between the current 891-person sample and the full population, we can identify any “Sampling bias” and see how the significance of factors like age and gender changes. This will allow us to verify our current findings hold true for the entire ship or if the missing data leads to different analytical conclusions.

<br>
<br>
