---
title: "Project Proposal"
output: html_document
date: "2026-04-07"
Authors: "Jayden Merrick, "
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# -- Project Proposal for 4/8/2026 -- #

# The names of the students in the group.

1. Jayden Merrick
2.Samantha Laderman
3.Dongwoon Han
4.Gavin Ryan
5.Hyewon Aha


# A brief description of how the group will communicate and work together.
# When during the week can the group meet? Consider needs of all group members.
# What platforms will the group use to communicate and share documents?

Group Planned Meeting (Time/Place): 3:00pm At Morgridge Tuesdays. Meet by Cafe.
Platform for Communication: iMessage for direct contact, Google docs for Code sharing, and Github for main Final Project file.


# The question or questions of interest that motivate the planned study.

Question: Does the higher the fare correlate with a higher survivability of the individuals in the dataset.


```{r}
# For the group: My data was put into a separate data folder
data <- read.csv("../data/titanic.csv")

head(data)
```


# A description of the data:

Who collected the data: Collected from Shanghai Jiao Tong University for Xiaming Chen's PhD studies.

How were the data collected: From blogs, answers, and user responses.

What are the key variables, what do they measure, and how are they related to your key questions: Fare and Survivability are the focus of our question. We should be able to reference the fare price to if the individual survived or not.

If the data is a sample from some larger population, what is that larger population: The data is a subet of the whole populaion. The actaul population would include the entire ~2,200 people that were on the titanic. Our data has around 800 individuals.
 

# Include a description of a graph which will show how the data informs the primary question of interest.

Describe what variables you will include and what type of graph you will create: 

 

# Finally, please do your best to identify which type of inference method you are going to use. In particular, please identify one of the inference methods listed below, and which variable(s) are your cause and effect (if applicable)

Method: Inference on a difference of two population proportions, as we deal with fare and survivability (dead/alive)
